Gammapy: A Python package for gamma-ray astronomy

### 2023-03-06 src/ms-review.tex

\abstract
	{
		Traditionally, TeV-\gammaray astronomy has been conducted
		by experiments employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array, will be operated as open observatories.
		Alongside the data, they will also make associated software tools available to a wider community.
    This necessity prompted the development of open, high-level astronomy software customised for high energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of its first long-term-support release, version 1.0.
		Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialised data products that
		are interoperable with other software packages.
	}
	{
		Starting from event list and instrument response functions,
		\gammapy provides the functionalities for reducing data binned in energy and sky coordinates.
		To handle the residual hadronic background, several techniques for background estimation
		are implemented in the package.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show the capabilities of \gammapy
		in multiple traditional and novel \gammaray analysis scenarios using public data
		such as spectral and spectro-morphological modelling and estimations of a spectral energy
		distribution and a light curve.
		Its flexibility and its power are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments:

\textit{Imaging Atmospheric Cherenkov Telescopes (IACTs)} obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by the cascading charged particles and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

\textit{Water Cherenkov Detectors (WCDs)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV and large duty-cycle, but a higher energy threshold and
usually lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

@@

%TODO: better transition into context
The Cherenkov Telescope Array (CTA) will be the first ground-based \gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by CTAO will be reconstructed event lists and corresponding instrument response functions.} will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of existing instruments and their interoperability,
it is required to use open data formats and open tools that can support the various analysis methods
commonly used in the field.

@@

In the sub-field of astronomy, it was the \astropy project \citep{astropy} that 
was created in 2012 to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units
as well as reading and writing FITS files.

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the idea of \astropy with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. Various \gammaray instruments
export their data to the standardised common \gadf data format. This data can then be combined and
analysed using a single common software library.
This means that the \gammapy package is not specific to any instrument,
but an independent community developed software project.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of gamma-ray sources as a function of the energy,
$E_{\rm true}$, and of the position in the FoV, $p_{\rm true}$:
\begin{equation}
    \Phi(p_{\rm true}, E_{\rm true}, \hat{\theta}) \quad \mathrm{[TeV^{-1}cm^{-2}s^{-1}]}
    \label{eq:general_source_model}
\end{equation}
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events, $N_{\rm pred}$, with their estimated energy $E$ and position $p$, the model
is convolved through the response function of the instrument.

@@

In total, the expected number of events in a \gammaray observation is given by:
\begin{equation}
	\begin{split}
  N(p, E, \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \circledast \left[ PSF \circledast \left( A_{\rm eff} \cdot t_{\rm obs} \cdot \Phi(\hat{\theta}) \right) \right]\\
                       & + Bkg(p, E) \cdot t_{\rm obs}
	\end{split}
\end{equation}

@@

Finally, predicted and observed events, $N_{obs}$, can be then combined in a likelihood function,
$\mathcal{L}(\hat{\theta}, N_{obs})$, usually Poissonian, that is maximised to obtain the best-fit parameters of the flux model, $\hat{\theta}$.

\subsection{Data analysis workflow}
The first step in \gammaray data analysis is the selection and extraction of observations
based of their metadata including information such as pointing direction, observation
time and observation conditions. The access to the events data and instrument 
reponse per observation is supported by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built-in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{o}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in different energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

@@

Figure~\ref{fig:data_flow} shows an overview of the different sub-packages and
their relation to each other. The \code{gammapy.data} and \code{gammapy.irf}
sub-packages define data objects to represent DL3 data, such as
event lists and IRFs as well as functionality
to read the DL3 data from disk into memory. The \code{gammapy.makers} sub-package
contains the functionality to reduce the DL3 data to binned maps.
Binned maps and datasets, which represent a collection of binned
maps, are defined in the \code{gammapy.maps} and \code{gammapy.datasets}
sub-packages, respectively. Parametric models, which are defined in
\code{gammapy.modeling}, are used to jointly model a combination
of datasets, for example, to make spectum using data from several facilities. Estimator classes,
which are contained in \code{gammapy.estimators}, are used to
compute higher level science products such as flux and signficance maps,
light curves or flux points. Finally there is a \code{gammapy.analysis}
sub-package which provides a high-level interface for executing analyses
defined from configuration files. In the following sections we will
introduce all sub-packages and their functionalities in more detail.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We also exclusively relied on publicly available
		data provided by the collaborations. The data is also availabe in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

The \code{gammapy.irf} sub-package contains all classes and functionality
to handle IRFs in a variety of formats.
Usually, \irfs store instrument properties in the form of multi-dimensional
tables, with quantities expressed in terms of energy (true or reconstructed),
off-axis angles or cartesian detector coordinates. The main information stored in
the common \gammaray \irfs are the effective area, energy dispersion,
PSF and background rate. The \code{gammapy.irf}
sub-package can open and access specific \irf extensions,
interpolate and evaluate the quantities of interest on both energy and spatial
axes, convert their format or units, plot or write them into
output files. In the following, we list the main classes of the
sub-package:

@@

%% CN: ref. for King function https://ui.adsabs.harvard.edu/abs/2011A%26A...534A..34R/abstract
%% CN: I would not say that the PSFKing takes into account "the gamma and sigma", fo what?
%% I would put a formula or quickly describe the function (e.g. it is a power-law, an exponential...)
\subsubsection{Point Spread Function}
\gammapy allows user to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussians (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three
Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing} takes into account the gamma and
sigma parameters. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
and offset from the pointing position.

@@

\subsubsection{Energy Dispersion}
For \iacts, the energy resolution and bias, or sometimes called energy dispersion,
is typically parametrised in terms of the so-called
migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. By definition, the mean of this ratio is
close to unity for a small energy bias and its distribution can 
be typically described by a Gaussian profile. However, more complex
shapes are also common. The migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are the \code{EnergyDispersion2D} which is
designed to handle the raw instrument description, and the \code{EDispKernelMap},
which contains an energy disperion matrix per sky position. I.e., a 4-dimensional
sky map where at each position is associated to an energy dispersion matrix.
The energy dispersion matrix is a representation of the energy resolution
as a function of the true energy only and implemnted in \gammapy
by the sub-class \code{EDispKernel}.

\subsubsection{Instrumental Background}
The instrumental background rate can be represented as either a 2-dimensional
data structure named \code{Background2D} or a 3-dimensional one named \code{Background3D}.
The background rate is stored as a differential count rate, normalised per solid angle
and energy at different reconstructed energies. In the first case the background is
expected to follow a radially symmetric shape and changes only with the offset angle from
FoV center. In the other case the background is allowed to vary with 
longitude and latitude of a tangential FoV coordinates system.

@@

The geometry object defines the pixelization scheme and map boundaries. It also
provides methods to transform between sky and pixel coordinates. Maps consist
of a geometry instance defining the coordinate system together with a
Numpy array containing the associated data. All map classes support a basic
set of arithmetic and boolean operations with  unit support, up- and downsampling
along extra axes, interpolation, resampling of extra axes, interactive visualisation
in notebooks and interpolation onto different geometries.

@@

\subsubsection{WCS Maps}
The FITS WCS pixelization supports a different number of projections to
represent celestial spherical coordinates in a regular rectangular grid.
\gammapy provides full support to data structures using this pixelization
scheme. For details see ~\cite{Calabretta2002}. This pixelisation
is typically used for smaller regions of interests, such as pointed
observations and is represented by a combination of the
\code{WcsGeom} and \code{WcsNDMap} class.

@@

\subsubsection{Region Maps}
In this case, instead of a fine spatial grid
dividing a rectangular sky region, the spatial dimension is reduced to a single
bin with an arbitrary shape, describing a region in the sky with that same
shape. Typically, they are used together with a non-spatial dimension, for
example an energy axis, to represent how a quantity varies in that dimension
inside the corresponding region. To avoid the complexity of handling
spherical geometry for regions, the regions are projected onto the local
tangential plane using a WCS transform. This approach follows Astropy's \textit{Regions}
package \citep{AstropyRegions2022}, which is both used as an API to define regions
for users as well as handling the underlying geometric operations. Region based
maps are represented by the \code{RegionGeom} and \code{RegionNDMap} classes.

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow access by name later in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_datasets}.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with the associated models and likelihood function, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistic, different types of Datasets are supported. The \code{MapDataset} is
used for combined spectral and morphological (3D) fitting, while a 1D spectral 
fitting can be performed using the \code{SpectrumDataset}.
While the default fit statistics for both of these classes is the \emph{Cash}~\citep{Cash}
statistic, there are other classes which support
analyses where the background is measured from control regions, so called \enquote{off} obervations.
Those require the use of a different fit statistics, which takes into account the
uncertainty of the background measurement. This case is covered by the \code{MapDatasetOnOff}
and \code{SpectrumDatasetOnOff} classes, which use the \emph{WStat}~\citep{WStat} statistic.

@@

Because the background models suffer from strong uncertainties it is required
to correct them from the data themselves. Several techniques are commonly used
in TeV \gammaray astronomy such as FoV background normalization or
background measurement in reflected regions, see~\cite{Berge07}.
Specific \code{Makers} such as the \code{FoVBackgroundMaker} or the
\code{ReflectedRegionsBackgroundMaker} are in charge of this process.
%% CN: I would add a single line explaining what each of them do

@@

When the expected number of background events is known, the statistic function 
is the so called \emph{Cash} statistic ~\citep{Cash}. It is used by datasets using background
templates such as the
\code{MapDataset}. When the number of background events is unknown and an \enquote{off}
measurement where only background events are expected is used, the statistic
function is \emph{WStat}. It is a profile log-likelihood statistic where the background
counts are marginalized parameters. It is used by datasets containing \enquote{off}
counts measurements such as the \code{SpectrumDatasetOnOff}, used for classical
spectral analysis.

@@

The \code{SkyModel} class represents the factorised model in Eq.~\ref{eq:source_model_dependency}
(the spatial and temporal components being optional).
A \code{SkyModel} object can represent the sum of several emission components:
either, for example, from multiple sources and from a diffuse emission, or from several spectral
components within the same source. To handle list of multiple \code{SkyModel} objects, \gammapy
implements a \code{Models} class.

%% CN: a link to the model gallery?
The model gallery provides a visual overview of the available models in
\gammapy. Most of the analytic models commonly used in \gammaray astronomy are
built-in. We also offer a wrapper to radiative models implemented in the Naima
package~\citep{naima}. The modeling framework can be easily extended with
user-defined models. For example, the radiaitve models of jetted Active Galactic Nuclei (AGN)
implemented in \agnpy, can be wrapped into \gammapy~\citep[see Section 3.5 of ][]{2022A&A...660A..18N}.

@@

\subsection{gammapy.visualization}
\label{ssec:gammapy-visualization}
The \code{gammapy.visualization} sub-package contains helper functions
for plotting and visualizing analysis results and \gammapy~data structures.
This includes for example the visualization of reflected background regions across
multiple observations or plotting large parameter correlation matrices of
\gammapy models. It also includes a helper class to split
wide field Galactic survey images across multiple panels to fit a standard
paper size.

>>>>

\abstract
	{
		Traditionally, TeV-\gammaray astronomy has been conducted
		by experiments employing proprietary data and analysis software.
		However, the next generation of \gammaray instruments,
		such as the the Cherenkov Telescope Array, will be operated as open observatories.
		Alongside the data, they will also make associated software tools available to a wider community.
    This necessity prompted the development of open, high-level astronomy software customised for high-energy astrophysics.
	}
	{
		In this article, we present \gammapy, an open-source Python package for the analysis of astronomical \gammaray data,
		and illustrate the functionalities of its first long-term-support release, version 1.0.
		Built on the modern Python scientific ecosystem, \gammapy provides a uniform platform for reducing and
		modelling data from different \gammaray instruments for many analysis scenarios.
    \gammapy complies with several well-established data conventions in high-energy astrophysics, providing serialised data products that
		are interoperable with other software packages.
	}
	{
		Starting from event lists and instrument response functions,
		\gammapy provides the functionalities for reducing data binned in energy and sky coordinates.
		Several techniques for background estimation are implemented in the package to handle the residual hadronic background.
		After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
		using Poisson maximum likelihood fitting
		and assuming a variety of spectral, temporal, and spatial models.
		Estimation of flux points, likelihood profiles and light curves is also supported.
	}
	{
		After describing the structure of the package, we show the capabilities of \gammapy
		in multiple traditional and novel \gammaray analysis scenarios using public data
		such as spectral and spectro-morphological modelling and estimations of a spectral energy
		distribution and a light curve.
		Its flexibility and its power are displayed in a final multi-instrument example,
		where datasets from different instruments, at different stages of data reduction,
		are simultaneously fitted with an astrophysical flux model.
	}{}

@@

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create large showers of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments:

\textit{Imaging Atmospheric Cherenkov Telescopes (IACTs)} obtain images of the atmospheric showers
by detecting the Cherenkov radiation emitted by charged particles in the cascade and
use these images to reconstruct the properties of the incident particle.
Those instruments have a limited field of view (FoV) and duty cycle, but
good energy and angular resolution.

\textit{Water Cherenkov Detectors (WCDs)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV and large duty-cycle, but a higher energy threshold and
lower signal-to-noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

@@

%TODO: better transition into context
The Cherenkov Telescope Array (CTA) will be the first ground-based \gammaray instrument to be operated as an open observatory.
Its high-level data\footnote{The lowest reduction level of data published by CTAO will be reconstructed event lists and corresponding instrument response functions.} will be shared publicly after
some proprietary period, and the software required to analyze it will be distributed
as well. To allow the re-usability of data from existing instruments and their interoperability,
it is required to use open data formats and open tools that can support the various analysis methods
commonly used in the field.

@@

In the sub-field of astronomy, the \astropy project \citep{astropy} was created in 2012
to build a community-developed core Python package for astronomy.
It offers basic functionalities that astronomers of many fields need, such as representing
and transforming astronomical coordinates, manipulating physical quantities including units
as well as reading and writing FITS files.

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the model of \astropy, with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. Various \gammaray instruments
export their data to the standardised common \gadf data format. This data can then be combined and
analysed using a common software library.
This means that the \gammapy package is not specific to any instrument,
but an independent community-developed software project.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

@@

To extract physically relevant information, such as the flux, spatial or spectral shape of one or more sources,
an analytical model is commonly adopted to describe the intensity of the radiation from gamma-ray sources
as a function of the energy, $E_{\rm true}$, and of the position in the FoV, $p_{\rm true}$:
\begin{equation}
    \Phi(p_{\rm true}, E_{\rm true}; \hat{\theta}) \quad \mathrm{[TeV^{-1}cm^{-2}s^{-1}]}
    \label{eq:general_source_model}
\end{equation}
where $\hat{\theta}$ is a set of model parameters that can be adjusted in a fit. To convert this analytical flux model
into a prediction on the number of gamma-ray events detected by an instrument, $N_{\rm pred}$,
with their estimated energy $E$ and position $p$, the model is convolved with the response function of the instrument.

@@

In total, the expected number of events in a \gammaray observation is given by:
\begin{equation}
	\begin{split}
  N(p, E; \hat{\theta})\ {\rm d}p\ {\rm d}E =  &E_{\rm disp} \circledast \left[ PSF \circledast \left( A_{\rm eff} \cdot t_{\rm obs} \cdot \Phi(\hat{\theta}) \right) \right]\\
                       & + Bkg(p, E) \cdot t_{\rm obs}
	\end{split}
\end{equation}

@@

Finally, predicted and observed events, $N_{obs}$, can be combined in a likelihood function,
$\mathcal{L}(\hat{\theta}, N_{obs})$, usually Poissonian, that is maximised to obtain the best-fit parameters of the flux model, $\hat{\theta}$.

\subsection{\gammapy data analysis workflow}
The first step in \gammaray data analysis is the selection and extraction of observations
based of their metadata including information such as pointing direction, observation
time and observation conditions. The access to the events data and instrument 
reponse per observation is supported by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. In addition to
providing a variety of built-in models, including spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky,
custom user-defined models are also supported.
Spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{0}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in different energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

@@

Figure~\ref{fig:data_flow} shows an overview of the different sub-packages and
their relation to each other. The \code{gammapy.data} and \code{gammapy.irf}
sub-packages define data objects to represent DL3 data, such as
event lists and IRFs as well as functionality
to read the DL3 data from disk into memory. The \code{gammapy.makers} sub-package
contains the functionality to reduce the DL3 data to binned maps.
Binned maps and datasets, which represent a collection of binned
maps, are defined in the \code{gammapy.maps} and \code{gammapy.datasets}
sub-packages, respectively. Parametric models, which are defined in
\code{gammapy.modeling}, are used to jointly model a combination
of datasets, for example, to compute a spectrum using data from several facilities. Estimator classes,
which are contained in \code{gammapy.estimators}, are used to
compute higher level science products such as flux and signficance maps,
light curves or flux points. Finally there is a \code{gammapy.analysis}
sub-package which provides a high-level interface for executing analyses
defined from configuration files. In the following sections we will
introduce all sub-packages and their functionalities in more detail.

@@

\subsection{gammapy.irf}
\label{ssec:gammapy-irf}
%
%
\begin{figure*}[ht!]
	\centering
	\includegraphics[width=1.\textwidth]{figures/irfs.pdf}
	\caption{
		Using \code{gammapy.irf} to read and plot instrument response functions.
		The left panel shows the effective area as a function of energy for
		the \cta, \hess, \magic, \hawc and \fermi instruments. The right panel shows
		the $68\%$ containment radius of the PSF as a function of energy for the \cta, \hess
		and \fermi instruments. The \cta IRFs are from the \textit{prod5} production. The \hess IRFs are from the DL3 DR1,
        using observation ID 033787. The \magic effective area is computed for a
        $20\,{\rm min}$ observation at the Crab Nebula coordinates. The
		\fermi IRFs use \textit{pass8} data and are also taken at the position of the Crab Nebula.
		The \hawc effective area is shown for the event classes $N_{Hit}=5 - 9$ as light gray
		lines along with the sum of all event classes as a black line. The \hawc IRFs are taken from
		the first public release of event data the \hawc collaboration. All IRFs do not correspond
		to the latest performance of the instruments, but still are representative of the 
		detector type and energy range. We exclusively relied on publicly available
		data provided by the collaborations. The data is also availabe in the
		\code{gammapy-data} repository.
    }
	\label{fig:irfs}
\end{figure*}
%

The \code{gammapy.irf} sub-package contains all classes and functionalities
to handle IRFs in a variety of formats.
Usually, \irfs store instrument properties in the form of multi-dimensional
tables, with quantities expressed in terms of energy (true or reconstructed),
off-axis angles or cartesian detector coordinates. The main quantities stored in
the common \gammaray \irfs are the effective area, energy dispersion,
PSF and background rate. The \code{gammapy.irf}
sub-package can open and access specific \irf extensions,
interpolate and evaluate the quantities of interest on both energy and spatial
axes, convert their format or units, plot or write them into
output files. In the following, we list the main classes of the
sub-package:

@@

%% CN: ref. for King function https://ui.adsabs.harvard.edu/abs/2011A%26A...534A..34R/abstract
%% CN: I would not say that the PSFKing takes into account "the gamma and sigma", fo what?
%% I would put a formula or quickly describe the function (e.g. it is a power-law, an exponential...)
\subsubsection{Point Spread Function}
\gammapy allows users to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussian distributions (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three
Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing} takes into account the gamma and
sigma parameters. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
and offset from the pointing position.

@@

\subsubsection{Energy Dispersion}
For \iacts, the energy resolution and bias, sometimes called energy dispersion,
is typically parametrised in terms of the so-called
migration parameter ($\mu$), which is defined as the ratio between the
reconstructed energy and the true energy. By definition, the mean of this ratio is
close to unity for a small energy bias and its distribution can 
be typically described by a Gaussian profile. However, more complex
shapes are also common. The migration parameter is given at each offset angle and
reconstructed energy. The main sub-classes are the \code{EnergyDispersion2D} which is
designed to handle the raw instrument description, and the \code{EDispKernelMap},
which contains an energy disperion matrix per sky position. I.e., a 4-dimensional
sky map where at each position is associated to an energy dispersion matrix.
The energy dispersion matrix is a representation of the energy resolution
as a function of the true energy only and implemnted in \gammapy
by the sub-class \code{EDispKernel}.

\subsubsection{Instrumental Background}
The instrumental background rate can be represented as either a 2-dimensional
data structure named \code{Background2D} or a 3-dimensional one named \code{Background3D}.
The background rate is stored as a differential count rate, normalised per solid angle
and energy interval at different reconstructed energies and offset angles.
In the \code{Background2D} case, the background is expected to follow a radially symmetric shape
and changes only with the offset angle from FoV center.
In the \code{Background3D} case, the background is allowed to vary with 
longitude and latitude of a tangential FoV coordinates system.

@@

The geometry object defines the pixelization scheme and map boundaries. It also
provides methods to transform between sky and pixel coordinates. Maps consist
of a geometry instance defining the coordinate system together with a
\numpy array containing the associated data. All map classes support a basic
set of arithmetic and boolean operations with  unit support, up and downsampling
along extra axes, interpolation, resampling of extra axes, interactive visualisation
in notebooks and interpolation onto different geometries.

@@

\subsubsection{WCS Maps}
The FITS WCS pixelization supports a number of different projections to
represent celestial spherical coordinates in a regular rectangular grid.
\gammapy provides full support to data structures using this pixelization
scheme. For details see ~\cite{Calabretta2002}. This pixelisation
is typically used for smaller regions of interests, such as pointed
observations and is represented by a combination of the
\code{WcsGeom} and \code{WcsNDMap} class.

@@

\subsubsection{Region Maps}
In this case, instead of a fine spatial grid
dividing a rectangular sky region, the spatial dimension is reduced to a single
bin with an arbitrary shape, describing a region in the sky with that same
shape. Region maps are typically used together with a non-spatial dimension, for
example an energy axis, to represent how a quantity varies in that dimension
inside the corresponding region. To avoid the complexity of handling
spherical geometry for regions, the regions are projected onto the local
tangential plane using a WCS transform. This approach follows Astropy's \textit{Regions}
package \citep{AstropyRegions2022}, which is both used as an API to define regions
for users as well as handling the underlying geometric operations. Region based
maps are represented by the \code{RegionGeom} and \code{RegionNDMap} classes.

\subsection{gammapy.datasets}
\label{ssec:gammapy-datasets}
%
\begin{figure}
	\small
	\import{code-examples/generated/}{gp_datasets}
	\caption{
        Using \code{gammapy.datasets} to read existing reduced binned datasets.
        After the different datasets are read from disk they are collected into a
        common \code{Datasets} container. All dataset types have an associated
        name attribute to allow a later access by name in the code. The
        environment variable \code{\$GAMMAPY\_DATA} is automtically resolved
        by \gammapy. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_datasets}.
    }
	\label{fig*:minted:gp_datasets}
\end{figure}
%
The \code{gammapy.datasets} subpackage contains classes to bundle
together binned data along with the associated models and likelihood function, which
provides an interface to the \code{Fit} class (Sec \ref{sssec:fit}) for
modeling and fitting purposes. Depending upon the type of analysis and the
associated statistic, different types of Datasets are supported. The \code{MapDataset} is
used for combined spectral and morphological (3D) fitting, while a 1D spectral 
fitting can be performed using the \code{SpectrumDataset}.
While the default fit statistics for both of these classes is the \emph{Cash}~\citep{Cash}
statistic, there are other classes which support
analyses where the background is measured from control regions, so called \enquote{off} obervations.
Those require the use of a different fit statistics, which takes into account the
uncertainty of the background measurement. This case is covered by the \code{MapDatasetOnOff}
and \code{SpectrumDatasetOnOff} classes, which use the \emph{WStat}~\citep{WStat} statistic.

@@

Because background models usually suffer from strong uncertainties, it is required
to correct them from the data themselves. Several techniques are commonly used
in TeV \gammaray astronomy such as FoV background normalization or
background measurement in reflected regions, see~\cite{Berge07}.
Specific \code{Makers} such as the \code{FoVBackgroundMaker} or the
\code{ReflectedRegionsBackgroundMaker} are in charge of this process.
%% CN: I would add a single line explaining what each of them do

@@

When the expected number of background events is known, the statistic function 
is the so called \emph{Cash} statistic ~\citep{Cash}. It is used by datasets using background
templates such as the
\code{MapDataset}. When the number of background events is unknown, and an \enquote{off}
measurement where only background events are expected is used, the statistic
function is \emph{WStat}. It is a profile log-likelihood statistic where the background
counts are marginalized parameters. It is used by datasets containing \enquote{off}
counts measurements such as the \code{SpectrumDatasetOnOff}, used for classical
spectral analysis.

@@

The \code{SkyModel} class represents the factorised model in Eq.~\ref{eq:source_model_dependency}
(the spatial and temporal components being optional).
A \code{SkyModel} object can represent the sum of several emission components:
either, for example, from multiple sources and from a diffuse emission, or from several spectral
components within the same source. To handle lists of multiple \code{SkyModel} objects, \gammapy
implements a \code{Models} class.

%% CN: a link to the model gallery?
The model gallery provides a visual overview of the available models in
\gammapy. Most of the analytic models commonly used in \gammaray astronomy are
built-in. We also offer a wrapper to radiative models implemented in the Naima
package~\citep{naima}. The modeling framework can be easily extended with
user-defined models. For example, the radiative models of jetted Active Galactic Nuclei (AGN)
implemented in \agnpy, can be wrapped into \gammapy~\citep[see Section 3.5 of ][]{2022A&A...660A..18N}.

@@

\subsection{gammapy.visualization}
\label{ssec:gammapy-visualization}
The \code{gammapy.visualization} sub-package contains helper functions
for plotting and visualizing analysis results and \gammapy~data structures.
This includes, for example, the visualization of reflected background regions across
multiple observations, or plotting large parameter correlation matrices of
\gammapy models. It also includes a helper class to split
wide field Galactic survey images across multiple panels to fit a standard
paper size.
